\documentclass[bigger]{beamer}

\input{header-beam} % change to header-handout for handouts

% ====================
\title[Lecture 19]{Logic I F13 Lecture 19}
\date{November 21, 2013}
% ====================

\include{header}

\setlength{\fitchprfwidth}{14em}

\section{Numerical Quantification}

\subsec{Numerical Quantification}{


\bit
\item Cardinal numbers can be determiners:
\bit
\item \emph{Three cubes} are left of \emph{two dodecahedra}
\eit
\item Not always clear if ``three cubes'' means \emph{exactly} or \emph{at least} three
\item We'll assume the latter---the ``exactly'' is implicated, not implied
\bit
\item Do you have two dollars? Yes, I have two dollars. (Uncontroversially true even if you have more than 2\$)
\item How much money do you have? I have two dollars. (True but misleading if you have more.)
\eit
\item FOL can express all of:
\bit
\item \emph{At least $n$} cubes are \dots 
\item \emph{Exactly $n$} cubes are \dots
\item \emph{At most $n$} cubes are \dots
\eit
\eit

}

\subsec{At least $n$}{

\bit
\item At least 1 cube is small:
\[
\exists x(Cube(x) \land Small(x))
\]
\item At least 2 cubes are small:
\[
\exists x\exists y(x \neq y \land Cube(x) \land Small(x) \land Cube(y) \land Small(y))
\]
\eit
}


\subsec{$\exists^{\ge n}x\, A(x)$}{

\bit
\item $\exists^{\ge n}x\, A(x)$:
\begin{align*}
\exists x_1\dots\exists x_n(x_1 \neq x_2 \land x_1 \neq x_3 \land \dots \land x_1 \neq x_n & {} \land {}\\
x_2 \neq x_3 \land \dots \land x_2 \neq x_n & {} \land {}\\
\ddots\qquad &\\
x_{n-1} \neq x_n & {} \land {}\\
 A(x_1) \land \dots \land A(x_n)) 
\end{align*}
\eit
}

\subsec{$\exists^{\ge n}x\, A(x)$}{


\bit
\item NB: must state that \emph{every pair} of variables is different, e.g.,
\[\sf\exists x_1\exists x_2\exists x_3(x_1 \neq x_2 \land x_2 \neq x_3 \land Cube(x_1) \land Cube(x_2) \land Cube(x_3))\]
only says ``There are at least two cubes''! 
\item At least $n$ As are Bs:\[
\exists^{\ge n} x(A(x) \land B(x))
\]
\eit
}

\subsec{Exactly One}{

\bit
\item There is exactly one cube:
\[
\exists x(Cube(x) \land \lnot \exists y(Cube(y) \land x \neq y))
\]
\item This is equivalent to:
\[
\exists x(Cube(x) \land \forall y(Cube(y) \to x = y))
\]
\item In general: ``x has property A \emph{uniquely}'':
\begin{align*}
A(x) \land & \forall y(A(y) \to x = y)\\
\text{or just:\qquad} & \forall y(A(y) \liff x = y)
\end{align*}
\eit
}

\subsec{$\exists^{=n} x\, A(x)$}{

\bit
\item There are exactly $n$ things that satisfy A:
\begin{align*}
\exists x_1\dots\exists x_n(x_1 \neq x_2 \land x_1 \neq x_3 \land \dots \land x_1 \neq x_n & {} \land {}\\
x_2 \neq x_3 \land \dots \land x_2 \neq x_n & {} \land {}\\
\ddots\qquad &\\
x_{n-1} \neq x_n & {} \land {}\\
\only<1>{ A(x_1) \land \dots \land A(x_n) &{} \land {}}\\
\forall y(A(y) \only<1>{\to}\only<2>{\leftrightarrow} (y = x_1 \lor \dots \lor y=x_n)))
\end{align*}
\item Exactly $n$ As are Bs:
\[
\exists^{=n}x (A(x) \land B(x))
\]
\eit

}

\subsec{At most $n$}{

\bit
\item There are \emph{at most $n$} As $\IFF$ There aren't \emph{at least $n+1$} As
\[
\exists^{\emph{\le n}}x\, A(x) \IFF \emph{\lnot}\exists^{\emph{\ge(n+1)}}x\, A(x)
\]
\item $\lnot\exists^{\ge(n+1)}x\, A(x)$ is equivalent to:
\begin{align*}
\forall x_1\dots\forall x_{n+1}((A(x_1) \land \dots \land A(x_{n+1})) & {} \to {}\\
(x_1 = x_2 \lor x_1 = x_3 \lor \dots \lor x_1 = x_{n+1} & {} \lor {}\\
x_2 = x_3 \lor \dots \lor x_2 = x_{n+1} & {} \lor {}\\
\ddots\qquad &\\
x_n = x_{n+1} & ))
\end{align*}
\eit
}

\subsec{Proofs With Numerical Claims}{

\fitchprf{
\pline{\exists x\, P(x)}\\
\pline{\forall x\forall y ((P(x) \land P(y)) \to x = y)}}{
\pline{\exists x(P(x) \land \forall y(P(y) \to x = y))}
}
}

\section{``The'', ``Both'', ``Neither''}

\subsec{Definite Descriptions}{

\bit
\item Definite description: \emph{the so-and-so}
\item Russell's analysis of definite description: to say\\
\bit
\item ``The A is B''\\
\eit
is to say:
\bit
\item There is a unique A
\item It is B
\eit
\item In FOL:
\[
\exists x(A(x) \land \forall y(A(y) \to x = y) \land B(x))
\]
\item or more succinctly:
\[
\exists x(\forall y(A(y) \liff x = y) \land B(x))
\]
\eit

}

\subsec{The vs. Exactly One}{

\bit \item Compare:
\ben
\item The cube is small
\[\exists x(Cube(x) \land \forall y(Cube(y) \to x = y) \land Small(x))\]
\item There is exactly one small cube
\[\exists x(Cube(x) \land \forall y((Cube(y) \emph{{}\land Small(y)}) \to x = y) \land Small(x))\]
\een
\item (2) can be true without (1), but not vice versa
\item So (2) is a FO consequence of (1), but not vice versa
\eit
}

\subsec{Strawson's Analysis}{

\bit
\item According to Russell, ``The cube is small'' is false if there is no cube, or if there is more than one
\item Strawson disagrees
\item ``There is a unique cube'' is not part of what is \emph{said}, but is only \emph{presupposed}
\eit

}

\subsec{``Both'' and ``Neither''}{

\bit
\item ``Both cubes are small'': There are exactly 2 cubes, and each of them is small:
\begin{align*}
\exists x\exists y(x \neq y \land Cube(x) \land Cube(y) &{} \land {}\\
\forall z(Cube(z) \to (z = x \lor z = y)) & {}\land {}\\
\forall z(Cube(z) \to Small(z))&)
\end{align*}
\item NB: ``Both cubes are small'' implies ``There are exactly two small cubes'', but not vice versa!
\item ``Neither cube is small'': There are exactly 2 cubes, and none of them are small:
\begin{align*}
\exists x\exists y(x \neq y \land Cube(x) \land Cube(y) & {}\land {}\\
\forall z(Cube(z) \to (z = x \lor z = y)) &{} \land {}\\
\forall z(Cube(z) \to \lnot Small(z))&)
\end{align*} 
\eit

}

\section{Truth-Functional Completeness}

\subsec{Truth Functions}{

\begin{block}{Definition}

An ($n$-place) \emph{truth function} $t$ is a mapping of $n$-tuples of \T{} and \F\ to either \T\ or \F.
\medskip

$n$-place truth functions correspond to truth tables of sentence with $n$ atomic sentences.
\medskip

A sentence $S$ containing the atomic sentences $A_1$, \dots, $A_n$
\emph{expresses} the truth function $t$ iff the truth value of $S$ on
the truth-value assignment which assigns $v_i$ to $A_i$ is $t(v_1,
\dots, v_n)$.
\medskip

An $n$-place truth function is \emph{expressible} if there is a
sentence containing atomic sentences $A_1$, \dots, $A_n$ that
expresses it.

\end{block}
}

\subsec{Examples}{

\begin{tabular}{cc@{\qquad}cc}
$\begin{array}{cc|c}
&& t_\land\\ \hline 
\T & \T & \T \\
\T & \F & \F \\
\F & \T & \F \\
\F & \F & \F
\end{array}$
&
$A_1 \land A_2$ &
$\begin{array}{cc|c}
&& t_\lor \\ \hline
\T & \T & \T \\
\T & \F & \T \\
\F & \T & \T \\
\F & \F & \F
\end{array}$
& $A_1 \lor A_2$
\end{tabular}
}

\subsec{Examples}{

\begin{tabular}{cc@{\qquad}cc}
$\begin{array}{cc|c}
& & t_1\\ \hline
\T & \T & \T \\
\T & \F & \T \\
\F & \T & \F \\
\F & \F & \F
\end{array}$
& $A_1$ &
$\begin{array}{cc|c}
& & t_{XOR}\\ \hline
\T & \T & \F \\
\T & \F & \T \\
\F & \T & \T \\
\F & \F & \F
\end{array}$
&
$(A_1 \lor A_2) \land \lnot(A_1 \land A_2)$ \\ \ \\
$\begin{array}{cc|c}
& & t_2\\ \hline
\T & \T & \T \\
\T & \F & \F \\
\F & \T & \T \\
\F & \F & \F
\end{array}$
& $A_2$ 
\end{tabular}
}



\subsec{Truth-Functional Completeness}{

\begin{block}{Definition}

A set of connectives is \emph{truth-functionally complete} if every
truth function is expressible by a sentence containing only those
connectives.

\end{block}

}

\subsec{$\{\land, \lor\}$ Not Truth-Functionally Complete}{

\bit
\item $\{\land, \lor\}$ is not truth-functionally complete
\item Remember: To be truth functionally complete, \emph{every} truth
  function would have to be expressible using only $\land$ and $\lor$
\item Which 2-place truth-functions can be expressed using $\land$ and
  $\lor$?
\item Not this one:
\[
\begin{array}{cc|c}
\T & \T & \F \\
\T & \F & \T \\
\F & \T & \T \\
\F & \F & \F
\end{array}
\]
  \eit

}


\subsec{Course Evaluation}{

\bit
\item Please fill out TWO TA evaluations, one for Jared and one for Dan.
\item More TopHat instant feedback questions? Or is TopHat not worth the trouble?
\item Three tests good, or would you prefer a midterm 
and a final after classes end?
\item Timing of homeworks---prefer Friday or Monday?
\item PASS sessions: did you go? Was it useful?
\eit
}
\end{document}

\subsec{$\{\land, \lor, \lnot\}$ is Truth-functionally Complete}{
\[
\begin{array}{lll|l@{\qquad}l}
A_1 & A_2 & A_3 & t_\mathrm{ODD} & S \\
\hline
\T & \T & \T & \T & \uncover<2->{(A_1 \land A_2 \land A_3)} \uncover<6->{{}\lor}\\
\T & \T & \F & \F \\
\T & \F & \T & \F \\
\T & \F & \F & \T & \uncover<3->{(A_1 \land \lnot A_2 \land \lnot A_3)} \uncover<6->{{}\lor}\\
\F & \T & \T & \F \\
\F & \T & \F & \T & \uncover<4->{(\lnot A_1 \land A_2 \land \lnot A_3)} \uncover<6->{{}\lor}\\
\F & \F & \T & \T & \uncover<5->{(\lnot A_1 \land \lnot A_2 \land A_3)} \uncover<6->{{}\lor}\\
\F & \F & \F & \F
\end{array}\]
}

\subsec{$\{\land, \lor, \lnot\}$ is Truth-Functionally Complete}{

\bit
\item $\lnot A_1 \land A_2 \land \lnot A_3$ is true in, and only in line \F\ \T\ \F
\item Combine using $\lor$: make $S$ true in all (and only) the lines where it is supposed to be true
\item NB: Procedure results in DNF!
\eit

}

\subsec{The ``neither\dots nor \dots'' connective: $\downarrow$}{

\[
\begin{array}{cc|c}
P & Q & (P \downarrow Q)\\
\hline
\T & \T & \F\\
\T & \F & \F\\
\F & \T & \F\\
\F & \F & \T
\end{array}
\]

}

\subsec{$\{\downarrow\}$ is Truth-Functionally Complete}{

\bit
\item Already know that $\{\lnot, \land, \lor\}$ is truth-functionally
  complete, i.e.,
\item Every truth-function can be expressed using only $\lor$, $\land$, $\lnot$
\item To show $\downarrow$ is truth-functionally complete, suffices to
  show that \emph{every sentence containing only $\lnot$, $\lor$, $\land$ is
  tautologically equivalent to one containing only $\downarrow$}
\item For that, it suffices to show that any negated sentence,
  conjunction, disjunction, can be expressed using only $\downarrow$
\eit

}

\subsec{Expressing $\lnot$ Using $\downarrow$}{

\begin{columns}
\begin{column}{3cm}
\[\begin{array}{cc|c}
P & Q & (P \downarrow Q)\\
\hline
\T & \T & \F\\
\T & \F & \F\\
\F & \T & \F\\
\F & \F & \T
\end{array}\]
\end{column}
\begin{column}{7cm}
\bits
\item Note how $P \downarrow Q$ is \F{} in the first line and \T{} in the last (when $P$ and $Q$ have same truth value)
\item So $P \downarrow P$ is \F{} if $P$ is \T, and \T{} if $P$ is \F, i.e., \[
\lnot P \Leftrightarrow (P \downarrow P)
\]
\eit
\end{column}
\end{columns}

}

\subsec{Expressing $\lor$ Using $\downarrow$}{

\begin{columns}
\begin{column}{3cm}
\[\begin{array}{cc|c}
P & Q & (P \downarrow Q)\\
\hline
\T & \T & \F\\
\T & \F & \F\\
\F & \T & \F\\
\F & \F & \T
\end{array}\]
\end{column}
\begin{column}{7cm}
\bits
\item $P \downarrow Q$ is the ``neither \dots nor'' connective, which can also
be expressed as $\lnot(P \lor Q)$, i.e.,
\[ \lnot(P \lor Q) \Leftrightarrow P \downarrow Q \] 
\item Negate both sides:
\[
P \lor Q \Leftrightarrow \lnot(P \downarrow Q)
\]
\item Apply what we figured out in last slide:
\[
P \lor Q \Leftrightarrow (P \downarrow Q)\downarrow(P \downarrow Q)
\]
\eit
\end{column}
\end{columns}

}

\subsec{Expressing $\land$ Using $\downarrow$}{

\begin{columns}
\begin{column}{3cm}
\[\begin{array}{cc|c}
P & Q & (P \downarrow Q)\\
\hline
\T & \T & \F\\
\T & \F & \F\\
\F & \T & \F\\
\F & \F & \T
\end{array}\]
\end{column}
\begin{column}{7cm}
\bits
\item $P \downarrow Q$ is the ``neither \dots nor'' connective, which can also
be expressed as $\lnot P \land \lnot Q$, i.e.,
\[ (\lnot P \land \lnot Q) \Leftrightarrow P \downarrow Q \] 
\item Equivalence holds for \emph{all sentences} $P$, $Q$, so also if we replace $P$ by $\lnot R$ and $Q$ by $\lnot S$:
\[
\lnot\lnot R \land \lnot\lnot S \Leftrightarrow (\lnot R \downarrow \lnot S)
\]
\item Delete $\lnot\lnot$'s, and express $\lnot$ using $\downarrow$:
\[
R \land S \Leftrightarrow (R \downarrow R)\downarrow(S \downarrow S)
\]
\eit
\end{column}
\end{columns}

}

\subsec{Truth-Functionally Complete Sets of Connectives}{

\bit
\item De Morgan's Law: $\land$ can be expressed by $\lor$ and $\lnot$
\item Similarly: $\lor$ can be expressed by $\land$, $\lnot$
\item So $\{\lor, \lnot\}$ and $\{\land, \lnot\}$ are
  truth-functionally complete
\item $\{\to, \bot\}$ is truth-functionally complete (HW)
\item $\{\to, \lnot\}$ is truth-functionally complete 
\item No other sets of connectives that don't contain one of these sets are complete 
\item ``Neither \dots nor'' is truth-functionally complete by itself
\item ``Not both'' connective is truth-functionally complete by itself (HW)
\item No other 2-place connectives are complete by themselves
\eit

}

\end{document}




